\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-28 papers/jfp/8-engineering.tex
Polishing S3-S4
More work tidying up S8

\subsection{Partial Stores and Exceptions}

@@

\item The \hs{restarting} scheduler can be made parallel in a few ways, but the
most direct is to have $n$ threads reading keys from the build queue. As before,
if a key requires a dependency not yet built, it is moved to the end~--~the
difference is that sometimes keys will be moved to the back of the queue not
because they are out of date but because of races with earlier tasks that had
not yet finished. As a consequence, over successive runs, potentially racey
dependencies will be separated, giving better parallelism over time.

\item The \hs{suspending} scheduler can be made parallel by starting static
dependencies of a \hs{Task} in parallel, while dynamic dependencies are being
resolved, using the strategy described by~\cite{marlow2014haxl}.
\end{enumerate}

@@

\todo{NM}{Should we add a paragraph or two giving a bit more details about how
parallelism is supported in \Shake?}

@@

\item[Non-determinism] Some tasks are \emph{non-deterministic}, producing a
result from a possible set. As an example, GHC when compiled using parallelism
can change the order in which unique variables are obtained from the supply,
producing different but semantically identical results.

@@

Overall we choose not to deal with cyclic dependencies, a choice that most build
systems also follow. There are computation frameworks that support tasks with
cyclic dependencies under the assumption that tasks are \emph{monotonic} in a
certain sense, e.g. see~\cite{pottier2009lazy} and~\cite{radul2009propagation}.

\subsection{Polymorphism and Multiple-Output Build Tasks}\label{sec-polymorphism}

@@

Generalised algebraic data types (GADTs)~\cite{spj2006gadts} can be used to
extend our models to support ``typed'' build tasks. The key idea is to replace
the monomorphic callback \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} by its
polymorphic equivalent \hs{fetch}~\hs{::}~\hs{k}~\hs{v}~\hs{->}~\hs{f}~\hs{v},
where~\hs{k}~\hs{v} is a GADT representing keys tagged by the type~\hs{v} of
corresponding values. The idea is best explained by way of an example:

@@

A more compelling example is build tasks that produce multiple output
keys~--~for example, \cmd{ghc A.hs} produces both \cmd{A.hi} and \cmd{A.o}.
That can be represented by having a key whose value is a pair of file names, and
whose result is a pair of file contents. From that, the task for \cmd{A.hi}
can be the first component of the result of the pair. Again, such an operation
can be encoded without polymorphic keys provided the pair of files (or a dummy
file representing the pair) is marked as changed if either of the contained
files change. Once again, polymorphic dependencies provide convenience rather
than power.

Note that multiple-output build tasks require one more convenience mechanism,
which is orthogonal to polymorphism. Multiple-output task descriptions partition
the set of keys into \emph{elementary subsets} e.g.
$\{\cmd{A.hi},\cmd{A.o}\} \rightarrow \cmd{ghcTask}$,
$\{\cmd{B.o}\} \rightarrow \cmd{gccTask}$, etc. It is very convenient to
allow to fetch arbitrary sets of keys, such as $\{\cmd{A.o},\cmd{B.o}\}$,
without manually decomposing them into appropriate elementary subsets. This
requires a method to efficiently construct task descriptions operating on
arbitrary sets of keys rather on elementary subsets. One way to implement such
``multiple-output task closure'' is shown below.

>>>>

\subsection{Partial Stores and Exceptions}\label{sec-failures}

@@

\item The \hs{restarting} scheduler can be made parallel in a few ways, but the
most direct is to have $n$ threads reading keys from the build queue. As before,
if a key requires a dependency not yet built, it is moved to the end~--~the
difference is that sometimes keys will be moved to the back of the queue not
because they are out of date but because of races with earlier tasks that had
not yet finished. As a consequence, if a the calcuation order is persisted over
successive runs (as \Excel does), potentially racey dependencies will be separated,
giving better parallelism over time.

\item The \hs{suspending} scheduler can be made parallel by starting multiple
dependencies simulataneously. One approach is to make the request for dependencies
take a list of keys, as implemented by \Shake. Another approach is to treat the
\hs{Applicative} dependencies of a \hs{Task Monad} in parallel, using the strategy
described by~\cite{marlow2014haxl}.
\end{enumerate}

Once sufficient parallelism is available the next challenge is preventing excess
parallelism and machine resource starvation, which is usually acheived with a thread pool or thread limit.

@@

@@

\item[Non-determinism] Some tasks are \emph{non-deterministic}, producing any
result from a possible set. As an example, GHC programs compiled using parallelism
can change the order in which unique variables are obtained from the supply,
producing different but semantically identical results.

@@

Overall we choose not to deal with cyclic dependencies -- a choice that most build
systems also follow. There are computation frameworks that support tasks with
cyclic dependencies under the assumption that tasks are \emph{monotonic} in a
certain sense, e.g. see~\cite{pottier2009lazy} and~\cite{radul2009propagation}.

\subsection{Polymorphic values}\label{sec-polymorphism}

@@

Generalised algebraic data types (GADTs)~\cite{spj2006gadts} can be used to
extend our models to support ``typed'' build tasks. The idea is to replace
the callback \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} by its
more polymorphic equivalent \hs{fetch}~\hs{::}~\hs{k}~\hs{v}~\hs{->}~\hs{f}~\hs{v},
where~\hs{k}~\hs{v} is a GADT representing keys tagged by the type~\hs{v} of
corresponding values. The idea is best explained by way of an example:

@@

\subsection{Multiple-Output Build Tasks}\label{sec-multiple-outputs}

We can use the extra polymorphism from \S\ref{sec-polymorphism} for build tasks that produce multiple output
keys~--~for example, \cmd{ghc A.hs} produces both \cmd{A.hi} and \cmd{A.o}.
That can be represented by having a key whose value is a pair of file names, and
whose result is a pair of file contents. From that, the task for \cmd{A.hi}
can be the first component of the result of the pair. Such an operation
can be encoded without polymorphic keys provided the pair of files (or a dummy
file representing the pair) is marked as changed if either of the contained
files change, but it is more elegant and type-safe with the extra polymorphism.

Multiple-output build tasks require one more convenience mechanism,
which is orthogonal to polymorphism. Multiple-output task descriptions partition
the set of keys into \emph{elementary subsets} e.g.
$\{\cmd{A.hi},\cmd{A.o}\} \rightarrow \cmd{ghcTask}$,
$\{\cmd{B.o}\} \rightarrow \cmd{gccTask}$, etc. It is very convenient to
allow to fetch arbitrary sets of keys, such as $\{\cmd{A.o},\cmd{B.o}\}$,
without manually decomposing them into appropriate elementary subsets. This
requires a method to efficiently construct task descriptions operating on
arbitrary sets of keys rather on elementary subsets. One way to implement such
``multiple-output task closure'' is shown below.
